
\section{General Functionality of EasyOCaml}

This section covers some general information on architecture of \easyocaml\ and
its language for orientation in supsequent parts.
It will give a broad overview on \easyocaml's course of actions first.
Every step is elaborated in more detail in subsequent sections, but they might
be easier to understand with the knowledge of their role in \easyocaml.
It will give a desription of the language supported by \easyocaml\ afterwards.

\subsection{Outline of EasyOCaml's Course}

By large \easyocaml's additions to the compiler and the toplevel loop do a
fairly similar job: First, they parse the command line flags. If they
contain the \texttt{-easy} flag, \easyocaml's type checker is enabled and flags
for defining language levels, teach packs and error printers are accepted and
loaden into \easyocaml.

Then, \easyocaml\ takes over the first stage of the compiler by parsing the
input with a \camlpf\ parser. The parser incorporates modifications by the
language level to the availabel syntax. The resulting AST is annotated with
fresh type variables.

\easyocaml\ then tries to type check the program. It reads the definitions in
the program and generates a set of constraints on the types of every element of
the program by traversing the AST. The constraints on the node's types are based
on the node's usage and their generation is detailed in the section
\ref{sec:typeinfer} and \ref{sec:rules}.
Then, the type checker tries to solve those constraints. If they are unifiable,
the type checker can assign a valid typing on the elements of the AST and gives
control back to \ocaml.
Otherwise, \easyocaml\ tries to generate as many conflict sets (type errors) as
possible from the constraints and reports them to the user.

\subsection{Supported Language: Caml$_{-m}$}
\label{sec:language}

\easyocaml\ targets to be usable not only for the very first steps in
programming, so a reasonable subset of the \ocaml\ language is supported.
This language can be characterized as ``Caml minus module declarations'',
hence its designation \camlm.

Unlike simpler functional programming languages like \textit{Scheme} with only
a single syntactic category \emph{expression}, \ocaml\ makes a distinction
between \emph{structure items}, \emph{expressions} and \emph{patterns}.  Here
are the supported language features in more detail, which also can be pruned by
teach packs. See section \ref{sec:grammar} for a complete grammar.


\subsection*{Structure items}

\ocaml\ programs consist of a list of \emph{structure items}, which are used to
declare values, types and exceptions.
\easyocaml\ supports
\begin{itemize}
  \item optionally parametrized \emph{type declarations} of type synonyms,
    records with optionally mutable fields and variant types.
  \item \emph{exception declarations} like those in \ocaml.
  \item optionally recursive (\texttt{rec}) and multiple (\texttt{and})
    \emph{value declarations}, where bindings occur with arbitrary patterns.
    Note, that a languge level may require type annotations for toplevel value
    declarations.
  \item Toplevel evaluations. Note that the toplevel evaluation of \texttt{e} is just
    syntactic sugar for the value declaration \texttt{let \_ = e}.
\end{itemize}

\subsection*{Core types}
\emph{core types} are direct combinations of existing types are called in \ocaml.
\easyocaml\ allows as core types primitive types (\texttt{int}, \texttt{char},
\texttt{string} and \texttt{float}), free (in type annotations) and bound
(within type declarations) type variables, type arrows (function types), tuples
and type constructors, i.e.\ applications of parametrized types.

\subsection*{Expressions}

Expressions are parts of a program which can be evaluated to a \ocaml\ value and
occur only as part of structure items.
\easyocaml\ supports simple expressions that can be found in \textsl{MiniML},
too, like variables, functions, infix operators, conditionals and variable
binding.

Despite those, it features the construction of tuples, records and variants,
conditionals with optional \texttt{else} branch, \texttt{while} and
\texttt{for} loops, sequences of expressions, exception handling (raising and
catching) as well as type annotations.

\subsection*{Patterns}

In \easyocaml, pattern matching is possible in every place where it works in
\ocaml, i.e.\ in value matching\ttfootnote{match \ldots\ with \emph{pat} ->
\ldots\ | \ldots}, in variable bindings\ttfootnote{let \emph{pat} = \ldots}, in
functional abstractions with the \texttt{function} keyword\ttfootnote{function
\emph{pat} -> \ldots\ | \ldots} and exception catching\ttfootnote{try \ldots\
with \emph{pat} -> \ldots\ | \ldots}.

Pattern matching works on every possible value in \easyocaml, i.e.\ primitive
values, tuples, variants and records\footnote{version 0.49 lacks the
implementation for pattern matching on the latter}, and can be nested at every
level.

